\MOC Server Pages:

### 2007-04-16 paper/paper.tex
svn-r24: many changes to paper; re-did unstaged versions of power and dir.
darcs-hash:20070416222907-e5a07-ca7ce5e89e572b776759e7697c439e40ad2e6db4.gz

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,xspace,graphicx,color}
\usepackage[leftno,lineno5]{lgrind}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml\xspace}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}

\newcounter{listing}
\setcounter{listing}{0}
\makeatletter
\def\ext@listing{lol}
\def\fnum@listing{Listing\nobreakspace\thelisting}
\makeatother
\newcommand{\mylst}[2]{%
  \captionof{listing}{\textbf{#1.} #2}%
  \label{lst:#1}%
  \lgrindfile{#1}}

\title{\MOC{} Server Pages:}
\subtitle{Web Publishing as Staged Computation}
\numberofauthors{1}
\author{
  \alignauthor Christopher League\\
  \affaddr{Long Island University · Computer Science}\\
  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
  \email{christopher.league@liu.edu}}
\conferenceinfo{\MOC{} Workshop}{October 2004, Vancouver}
\CopyrightYear{2004}
\toappear{Submitted to the \MOC{} Workshop (Vancouver,
  2004).}

\begin{document}
\maketitle
\begin{abstract}
  Modern dynamic web services are really computer programs.
  Some parts of these programs run off-line, others run
  server-side on each request, and still others run within
  the browser.  In other words, web publishing is
  \emph{staged computation,} either for better performance,
  or because certain resources are available in one stage
  but not another.  Unfortunately, the various web
  programming languages target one stage only.  This is a
  tremendous opportunity for multi-stage languages in
  general, and for \MOC{} in particular.

  We present the design of \MOC{} Server Pages.  Unlike
  other languages in its genre, the embedded \MOC{} code
  blocks may be annotated with staging information, so that
  the programmer may safely and precisely control which
  computation occurs in which stage.  A prototype web
  server, written in OCaml, supports web sites with both
  static and dynamic content.  We provide several sample
  programs and demonstrate the performance gains won using
  multi-stage programming.
\end{abstract}
\category{D.3.2}{Programming Languages}{Language
  Classifications}[Specialized application languages]
\category{H.3.5}{Information Storage and Retrieval}{Online Information
  Services}[Web-based services]

\vfill
\section{Motivation}
\label{sec:intro}

@@

We present ``\MOC{} server pages,'' a new domain-specific
language for web applications programming.  It leverages the
staging annotations of \MOC~\cite{calcagno03meta,taha00metaml}
to provide safe and precise control over the each stage of the
computation.  We have implemented the system in two parts: our
translator transforms the server page language into a \MOC{}
module, which then can be incorporated into our multi-threaded
HTTP 1.1 server (also written in \MOC{}).  The scalability
gained by staging certain applications is stunning.

The next section sketches the design and translation of \MOC{}
server pages, and section~\ref{sec:eg} includes some non-trivial
examples.  The server implementation is decribed in
section~\ref{sec:impl}.  Performance and scalability are
discussed in section~\ref{sec:perf}

@@

The general idea of a \emph{server page} language is that we
  write HTML by default, and embed code %
\cd{\Mquote{}like this\Munquote}.
PHP programmers are familiar with this syntax for embedding code, but
in our case, the code itself is written in \MOC.  Here is your first
\MOC{} server page:
\lgrindfile{first.meta}
and its output:
\begin{quote}
  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
\end{quote}
If a code block raises an exception, the message is sent to the
client's browser in bold-face, and the rest of the page is
aborted.

A very common use of code blocks is to print out (i.e., send to
the browser) the result of evaluating some expression.  The
syntax \cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must
have type \cd{string}.  Alternatively, messages may be formatted
with \cd{printf} by placing the format string immediately after
the code delimiter: \lgrindfile{pi.meta}
\begin{quote}
  ........ $\pi$ ÷ 008 is 0.3927
\end{quote}
In a departure from the standard OCaml syntax for \cd{printf}, the
arguments are comma-delimited.
%%
One more kind of code delimiter is used for \emph{declarations;}
these are evaluated during the publishing stage:
\lgrindfile{perm.meta}
\begin{quote}
Permissions on current directory are 0755
\end{quote}

@@

To see how all this works, let's consider how a \MOC{} server
page is translated into a proper \MOC{} program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
Listing~\ref{lst:trans.meta} shows a sample \MOC{} server page,
and listing~\ref{lst:trans.ml} contains its translation.

@@

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins on line~{???line:page} of
listing~\ref{lst:trans.ml}.  It is a function of two arguments:
\cd{req} encapsulates the HTTP request, including the headers
and query arguments; \cd{puts} is a function with type
\cd{string\to unit}; it transmits text across the network to the
visitor's browser.  \cd{arg} looks up an
argument in the query string of the request; its type is
\cd{string\to string option}.  These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC{} type system
prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically follows
code blocks with either a semi-colon or the {\KWfont in}
keyword, as needed.

Now that we understand how the server page is assembled into a
staged program, the effects of adding \MOC{} staging operators
to our pages should be predictable.  Below is a reformulation of
the permissions example, using escape \Msplice{} and \cd{lift} to
ensure that \cd{stat} is called only once.
\lgrindfile{perm2.meta}
%%
Still, the \cd{printf} conversion is performed at serve time, so
maybe it is best to pre-generate that as well:
\lgrindfile{perm3.meta}
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  Finally it is possible to splice in
additional code at publish time, such as in this count-down
example:
\lgrindfile{count.meta}
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

These tiny examples illustrate some of the common uses of the
escape operator, for which we developed syntactic sugar:

@@

One last feature in the design of \MOC{} server pages is the ability
to take publish-stage arguments.  The parameter names are specified to
the translator in a \emph{pragma} declaration, which can appear
anywhere in the code:

@@

Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC{} code is, for now, an implementation detail
(see section~\ref{sec:impl}).

@@

\lgrindfile{navspec-sig.ml}

The first function generates the standard XHTML header given the
title, while the second function generates the navigation bar given
the URI denoting the current page.  These can be called at either
stage.  Naturally, if the title includes a dynamic argument, then
\cd{preamble} will be delayed until the serve stage.

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \cd{Num} module.  The implementation of this
service appears in listing~\ref{lst:power.meta}.

@@

The first function wraps a string into a fixed width block.  The power
function itself begins on line~{???line:power-fun}.  No output should
occur until the preamble is generated on
line~{???line:power-preamble}.  The user's input is translated to a
number relatively late (line~{???line:power-num}); if there is a
problem here, the navigation bar and form will have already been
output before the exception message.

In section~\ref{sec:perf}, we will measure the impact of staging on
the scalability of this service.  To derive the un-staged version for
comparison, we simply remove all the brackets and escapes from the
code in listing~\ref{lst:power.meta}.

@@

Many web servers can be configured to permit clients to browse
directories.  The web server generates, on the fly, an HTML page
containing the names and attributes of (and hyperlinks to) all
the files in the directory.  In Apache, the
\texttt{mod\_autoindex} module provides this feature.
%%
ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
complex example of the same idea. It is a Python program that sits on
the web server and allows remote users to browse a CVS repository with
a standard web browser.

This kind of service can be fairly intense; one HTTP request is likely
to generate dozens of system calls and disk accesses.  If the
directory is viewed more often that it is changed, then it makes sense
to cache or pre-generate the pages.  Although it is simple enough to
write a script to generate directory pages off-line, what if we also
want dynamic user-controlled behavior, such as sorting and filtering?

We implemented
contribution to the genre not only lists files in a given
directory, but displays their MD5 checksums, renders their sizes in
human-readable form (`1.2M' rather than `1194822'), and colorizes their
names based on their extension or kind.  At serve time, the user may
specify a regular expression for filtering, and one of 5 criteria for
sorting.  The screen shot in figure~\ref{fig:server-dir} shows the
result of browsing the source directory of the server.

@@

This service is a bit like the conference calendar outlined in
section~\ref{sec:intro}.  Everything that can be pre-generated
\emph{is.}  At serve time, we just compile the regular expression,
then filter and sort the data (unless the default order was chosen).
The text of each row is ready and waiting.  At serve time, the files
need not be opened or even \id{stat}ted.  The crux of the
implementation is shown in listing~\ref{lst:dir.meta}:

@@

In the first stage, this reads all the filenames in the directory and
sorts them by name.  Then it looks up the fileinfo, computes the MD5
sum, and prepares the text of each row.  Finally, it generates code
that matches the filename against a (future-stage) regular expression,
and includes it in a list only if it matches.  In the next stage, this
list will be constructed and sorted (if necessary).

@@

Currently, the \MOC{} implementation is not capable of reading
and writing \cd{code} values to a file.  This means that
generated server pages (after the publish stage) must already be
in the server memory when it starts.  This is an unfortunate
limitation, as it means we must restart the entire web server to
re-publish any particular page!

@@

\section{Performance}
\label{sec:perf}

One of the benefits of a \emph{compiled} implementation of staged
programming---even byte-code compiled, like \MOC{}---is that it
permits meaningful measurement.  Here, we describe the performance
characteristics of the prototype, focusing in particular on the
benefits of staging the computations performed by various web
services.

The single most important datum for evaluating web server performance
is \textit{throughput:} the number of requests successfully answered
per unit of time.  To establish a baseline, we first tested the
throughput of the Caml prototype HTTP server delivering chunks of
static data of various sizes, up to 64k bytes.
%%
Measurements were taken on an otherwise idle 1.8GHz Intel
Xeon\texttrademark{} workstation\footnote{with 512kB cache, 768MB RAM,
  and Ultra160 SCSI} running Linux 2.4.  We used \texttt{ab},
the Apache HTTP server benchmarking tool, to issue requests
from 8 threads simultaneously for 30 seconds.\footnote{\texttt{ab}
  version 1.3d, available from \url{http://httpd.apache.org/}, 
  invoked like this:
  \texttt{ab -k -t 30 -c 8 }\textit{url}}

@@

The baseline results are shown in figure~\ref{fig:static}.  The
series marked “Caml file” was served by the \texttt{FileHandler}
module.  For the “Caml code” series, the same files were treated
as \MOC{} Server Pages, and thus turned into one big
\texttt{puts} statement, executed by the \texttt{CodeHandler}
module.  The only reason the code wins in the beginning is that
all code pages are already loaded into the server’s memory on
startup (to work around a limitation of the prototype---see
sections~\ref{sec:impl} and~\ref{sec:future}), but the
\texttt{FileHandler} must read files from disk each time.

Figure~\ref{fig:static} includes comparable results for PHP, the
popular server-side computation system.\footnote{libphp4.so
  (version 4.3.4) loaded into Apache 1.3} Here, we gave the same
static data files the extension \texttt{.php}, so that Apache
would treat them as PHP code, even though they have no
\cd{\Mquote{}code blocks\Munquote}.
%%
We omitted the results for Apache serving static files, because they
are way off scale: Apache handled an astounding 4,438 hits per second
for the 1k file, and 1,813 for the 64k file.  Apache is heavily
optimized for serving static files: apart from caching, it uses the
special \texttt{sendfile()} system call for zero-copy file transfer
from kernel space~\cite{tranter03sendfile}.  Although our prototype is
no match for Apache on static files, the overhead for interpreting
code seems no worse than that of PHP.
%%

@@

Finally, we look at the performance of staged and unstaged
directory browsing; see figure~\ref{fig:browse}.  For this task,
we created directories containing a fixed number of files with
random data.  The average file size was 32k.  The \emph{x}~axis
shows the number of files in the directory.

@@

Directory browsing was the most realistic of the examples, and
the benefit of staging is crystal clear.  In browsing a
directory of 64 files, the unstaged program barely answered 18
requests per second.  It would certainly succumb to the Slashdot
effect!

This is exactly the kind of page where the real work needs to be
done in advance.  But that doesn't mean it needs to be a
completely static page, either.  By carefully staging the
computation, we gathered the file information in advance, yet
still filtered and sorted the results on demand.  The staged
directory browser answered more than 550 requests (for the same
64-file directory) per second.

\section{Related work}
\label{sec:related}

We already referenced the other systems for server-side
computation, but we would be remiss without mentioning
XCaml~\cite{baretta04xcaml}\ldots

For client-side computation, besides JavaScript and Java
applets, there is the work by \citet{rouaix96web} on Caml
applets.  It may be possible to update his MMM browser for use
with \MOC{}, but we would prefer to use a plug-in approach that
works with mainstream browsers\ldots

\citet{elsman04typing}, \citet{wallace99haxml},
\citet{hosoya03xduce}, and \citet{ohl04xhtml} validate HTML code
in functional languages.  We could probably use OCaml's type
system to greater advantage this way, but it may require
validating the text parts of the server page with a more
mainstream SGML or XML validator.

\section{Future directions}
\label{sec:future}

 This really needs to be a module for a real server, such
  as Apache or AOLserver.

 Need more serious case studies, where one logical service
  is implemented across a series of pages.

 Need to address the major implementation shortcoming, of
  loading all the generated pages into memory on startup.

  Marshalling closures works partly in OCaml, but only if they
  are part of the ``original program''---that's probably too
  limiting for our case.  The abstract syntax representation of
  code values could be dumped to a file, but the `persistent'
  functions it refers to will need to be stored somewhere\ldots

>>>>

\documentclass[preprint]{acm_proc_article-sp}
\usepackage{ucs,textcomp,url,xspace,graphicx,color}
\usepackage[nolineno]{lgrind}
\usepackage[bf]{caption2}
\usepackage[numbers]{natbib}
\usepackage[utf8]{inputenc}
\ifx\pdfoutput\undefined\else
  \usepackage[
    pdfpagemode=None,
    pdftitle={MetaOCaml Server Pages},
    pdfsubject={Web Publishing as Staged Computation},
    pdfauthor={Christopher League}
  ]{hyperref}
\fi
\input{fonts}
\def\MOC{MetaOCaml\xspace}
\hyphenation{Meta-OCaml data-base Slash-dot}
\def\quote{\list{}{\leftmargin1.4em\rightmargin1em}\item[]}
\newcommand{\myfig}[3]{%
  \begin{figure}[tbp]%
    #3%
    \caption{#2}%
    \label{fig:#1}%
  \end{figure}}
\newcommand{\screenshot}[2]{%
  \myfig{#1}{#2}
  {\centering%
    \includegraphics[width=20pc,keepaspectratio]{#1}}}
\newcommand{\gnuplot}[2]{%
  \myfig{#1}{#2}{\input{#1}}}
\newcommand{\listing}[2]{\lagrind[tbp]{#1}{#2}{lst:#1}}

\newcounter{listing}
\setcounter{listing}{0}
\makeatletter
\def\ext@listing{lol}
\def\fnum@listing{Listing\nobreakspace\thelisting}
\makeatother
\newcommand{\mylst}[2]{%
  \captionof{listing}{\textbf{#1.} #2}%
  \label{lst:#1}%
  \lgrindfile{#1}}
\newcommand{\mylstf}[2]{%
  \begin{figure}[tbp!]%
    \lgrindfile{#1}%
    \captionof{listing}{\textbf{#1.} #2}%
    \label{lst:#1}%
  \end{figure}}

\title{\MOC Server Pages:}
\subtitle{Web Publishing as Staged Computation}
\numberofauthors{1}
\author{
  \alignauthor Christopher League\\
  \affaddr{Long Island University · Computer Science}\\
  \affaddr{1 University Plaza · Brooklyn, NY 11201}\\
  \email{christopher.league@liu.edu}}
%\conferenceinfo{\MOC Workshop}{October 2004, Vancouver}
%\CopyrightYear{2004}
%\toappear{Submitted to the \MOC Workshop (Vancouver,
%  2004).}

\begin{document}
\maketitle
\begin{abstract}
  Modern dynamic web services are really computer programs.
  Some parts of these programs run off-line, others run
  server-side on each request, and still others run within
  the browser.  In other words, web publishing is
  \emph{staged computation,} either for better performance,
  or because certain resources are available in one stage
  but not another.  Unfortunately, the various web
  programming languages target one stage only.  This is a
  tremendous opportunity for multi-stage languages in
  general, and for \MOC in particular.

  We present the design of \MOC Server Pages.  Unlike
  other languages in its genre, the embedded \MOC code
  blocks may be annotated with staging information, so that
  the programmer may safely and precisely control which
  computation occurs in which stage.  A prototype web
  server, written in OCaml, supports web sites with both
  static and dynamic content.  We provide several sample
  programs and demonstrate the performance gains won using
  multi-stage programming.
\end{abstract}
\category{D.3.2}{Programming Languages}{Language
  Classifications}[Specialized application languages]
\category{H.3.5}{Information Storage and Retrieval}{Online Information
  Services}[Web-based services]

\section{Motivation}
\label{sec:intro}

@@

We present ``\MOC server pages,'' a new domain-specific
language for web applications programming.  It leverages the
staging annotations of \MOC~\cite{calcagno03meta,taha00metaml}
to provide safe and precise control over the each stage of the
computation.  We have implemented the system in two parts: our
translator transforms the server page language into a \MOC
module, which then can be incorporated into our multi-threaded
HTTP 1.1 server (also written in \MOC).  The scalability
gained by staging certain applications is stunning.

The next section sketches the design and translation of \MOC
server pages, and section~\ref{sec:eg} includes some non-trivial
examples.  The server implementation is decribed in
section~\ref{sec:impl}.  Performance and scalability are
discussed in section~\ref{sec:perf}

@@

The general idea of a \emph{server page} language is that we
  write HTML by default, and embed code %
\cd{\Mquote{}like this\Munquote}.
PHP programmers are familiar with this syntax for embedding code, but
in our case, the code itself is written in \MOC.  Here is your first
\MOC server page:
\lgrindfile{first.meta}
and its output:
\begin{quote}
  \textit{Hello,} \textbf{Unhandled exception: Failure("Nice try!")}
\end{quote}
If a code block raises an exception, the message is sent to the
client's browser in bold-face, and the rest of the page is
aborted.

A very common use of code blocks is to print out (i.e., send to
the browser) the result of evaluating some expression.  The
syntax \cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must
have type \cd{string}.  Alternatively, messages may be formatted
with \cd{printf} by placing the format string immediately after
the code delimiter: \lgrindfile{pi.meta}
\begin{quote}
  ........ $\pi$ ÷ 008 is 0.3927
\end{quote}
In a departure from the standard OCaml syntax for \cd{printf}, the
arguments are comma-delimited.  This way, fewer parentheses are
required when using the escape or lift operators on the arguments.

One more kind of code delimiter is used for \emph{declarations;}
these are lifted to the top of your program, and evaluated during
the publishing stage:
%%
\lgrindfile{perm.meta}
\begin{quote}
Permissions on current directory are 0755
\end{quote}

@@

\subsection{Review of staging annotations}

\MOC augments OCaml with just three annotations, to indicate how
programs are to be staged.  Brackets \cd{\Mbegincode like
  these\Mendcode} construct future-stage computation.  The code within
is not executed \emph{now,} but just returned, as a code value that can be
run later.

Within brackets, the splice or escape operator \cd{\Msplice e} may
appear.  It interrupts the code construction to execute the expression
\cd{e} (in the current stage) and splice its result into the
future-stage computation.  Thus, \cd{e} is required to return code of
the proper type.

To compute and splice in a regular (non-code) value, we define a
function \cd{{\KWfont let} lift x = \Mbegincode x\Mendcode}; this
takes any value and turns it into code.  We use it like this:
\cd{\Mbegincode 2 * \Msplice(lift(3+4))\Mendcode} The addition is
performed immediately (because it is escaped), and the result is
spliced into the code, producing \cd{\Mbegincode 2 * 7\Mendcode}.

Finally, there is an operator \Mrun{} to execute constructed code.
Applying it to the example above, \cd{\Mrun{} \Mbegincode 2 * 7
  \Mendcode} produces \cd{14}.

\subsection{Translation to \MOC}
To see how all this works, let's consider how a \MOC server
page is translated into a proper \MOC program, to be executed
at publish time.  Since the program is executed before any
browser has requested the page, it cannot directly return or
output HTML.  Instead, it will construct and return a \cd{code}
object which is subsequently run on each request (serve stage).
Listing~\ref{lst:trans.meta} shows a sample \MOC server page,
and listing~\ref{lst:trans.ml} contains its translation.

@@

Declaration blocks have been lifted to the top; any side effects
contained there are executed when the page is published.  The
constructed code begins with the \cd{{\KWfont let} page} line in
listing~\ref{lst:trans.ml}.

The \cd{a b c} represent publish-stage arguments (the names are
specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to 
transmit text across the network to the user's browser.  \cd{arg} is
then defined to look up an argument in the query string of the
request; its type is \cd{string\to string option}.

These are all \emph{serve} (second)
stage values.  It makes no sense to access them in the
\emph{publish} (first) stage, and indeed the \MOC type system
prevents this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
listing~\ref{lst:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically follows
code blocks with either a semi-colon or the {\KWfont in}
keyword, as required.

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is a reformulation of the permissions
example, using escape \Msplice{} and \cd{lift} to ensure that
\cd{stat} is called only once.%
\lgrindfile{perm2.meta}%
Still, the \cd{printf} conversion is performed at serve time, so maybe
it is best to pre-generate that as well:%
\lgrindfile{perm3.meta}%
Now, all that remains to execute at serve time is %
\cd{puts "0755"}.  Finally it is possible to splice in serve-stage
parameters at publish time, such as in this count-down example:%
\lgrindfile{count.meta}%
It removes the loop overhead by expanding to a sequence of 99
\cd{puts} statements.

These tiny examples illustrate some of the common uses of the
escape operator, for which we developed syntactic sugar.  The first
set of code blocks we define use the tilde character to indicate that
some part of the enclosed quote is escaped:

@@

It is also common to use the escape with \cd{lift}; this means we are
computing a value immediately and splicing it into the code.  For
these blocks, we use the \Mbang{} character:

\begin{tabular}{{l@{\quad$\leadsto$\quad}l}}
\cd{\Mquote{\Mbang} a \Munquote} &
\cd{\Mquote{} \Msplice(lift( a )) \Munquote}
\\
\cd{\Mquote{\Mbang=} b \Munquote} &
\cd{\Mquote{=} \Msplice(lift( b )) \Munquote}
\\
\cd{\Mquote{\Mbang}let x = c \Munquote} &
\cd{\Mquote{} let x = \Msplice(lift( c )) \Munquote}
\\
\cd{\Mquote{\Mbang}\id{"fmt"} d, e\Munquote} &
\cd{\Mquote{}\id{"fmt"} \Msplice(lift(d)), \Msplice(lift(e))\Munquote}
\\
\end{tabular}

We will see examples of most of these blocks in section~\ref{sec:eg}.

\subsection{Publish-stage arguments}
One last feature in the design of \MOC server pages is the ability
to take publish-stage arguments.  The parameter names are specified to
the translator in a \emph{pragma} declaration, which can appear
anywhere in the code:

@@

Now the code can be instantiated with early parameters in countless
ways.  For example, requesting the URI \id{/longcount} could print a
count-down from 99, while \id{/shortcount} starts at 9 instead.  Both
are generated from the same server page.  Mapping from URIs to
instantiated \MOC code is, for now, an implementation detail
(see section~\ref{sec:impl}).

@@

\lgrindfile{navspec-sig.ml}
%%
The functions above may be invoked at any stage.  Naturally, if the
page title includes a dynamic argument, then \cd{preamble} will have
to be delayed until theh serve stage.

@@

@@

Judging from its prevalence in the literature, we are certain that
millions of grateful users would subscribe to an online service
capable of computing the \emph{power} function!  The screen shot in
figure~\ref{fig:power127} illustrates how it works.  The navigation
bar shows the exponents for which code has been
pre-generated.\footnote{In the current implementation, it is not
  possible for the user to request other exponents once the server is
  running.  See section~\ref{sec:impl} for an explanation.}  After
selecting the exponent, the user types the base into the form and
presses return.  The result is computed using OCaml's
arbitrary-precision \cd{Num} module.  The complete, documented script
appears in listing~\ref{lst:power.meta}.

The user's input is converted to a number relatively late in the
script.  If \cd{num\_of\_string} generates an exception (perhaps
because the user typed non-numeric text into the box), the navigation
bar and form will have already been output before the error message
appears.

In section~\ref{sec:perf}, we will measure the impact of staging on
the scalability of this application.  To derive the un-staged version
for comparison, we simply convert \cd{\Mquote{\Mtilde}} and
\cd{\Mquote{!}} blocks into plain \cd{\Mquote{}} and remove all other
brackets and escapes from the code in listing~\ref{lst:power.meta}.

@@

Now we consider a more substantial example.
Many web servers can be configured to permit clients to browse
directories.  The web server generates, on the fly, an HTML page
containing the names and attributes of (and hyperlinks to) all
the files in the directory.  In Apache, the
\texttt{mod\_autoindex} module provides this feature.
%%
ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
complex example of the same idea; it allows remote users to browse a
CVS repository with a standard web browser.

This kind of service can be fairly intense; one HTTP request is likely
to generate dozens of system calls and disk accesses.  If the
directory is viewed more often that it is changed, then it makes sense
to cache or pre-generate the pages.  Although it is simple enough to
write a script to generate directory pages off-line, what if we also
want dynamic behavior, such as user-controlled sorting and filtering?

Our implementation not only lists files in a given directory, but
displays their MD5 checksums, renders their sizes in human-readable
form (`1.2M' rather than `1194822'), and colors their names based on
their extension or file type.  We also support dynamic (serve time)
customization: the user may specify a regular expression for
filtering, and select one of 5 criteria for sorting.  The screen shot
in figure~\ref{fig:server-dir} shows the result of browsing the source
directory of the server itself.

@@

This service is a bit like the conference calendar proposed in
section~\ref{sec:intro}.  We pre-generate \emph{everything} that does
not rely on serve-time parameters.  The stats and MD5 sums of the
files are collected \emph{once,} in the publish stage.  The text of
each possible row is prepared in advance.  Once the user's request is
made, we filter and sort the list, then output the pre-generated text
of each remaining row.  The files need not be opened or even
\id{stat(2)}ed while the user is waiting.  Listing~\ref{lst:dir.meta}
contains the complete script, with documentation.

@@

Much of the above code is unaffected by staging; functions like
\cd{sort\_by}, \cd{human\_size}, and even \cd{fileinfo} are oblivious
to the stage in which they are run.  Therefore, the staged code comprises a
relatively small portion of the entire program: mainly the function
\cd{list\_files} and its call site.

To understand the staging technique used here, it may help to examine
the output of just the first stage of computation.  The following
listing is cleaned up from the \MOC pretty-printer, with references to
compiled values resolved in-place.

\lgrindfile{dir-out.ml}

The \cd{list} is formed by testing filenames (in reverse alphabetical
order) against the compiled regular expression \cd{rc}, including only
those that match.  The file info record for each file has already
been prepared and appears in the code as a value.

Because the file information is compiled into the code, any changes to
the file system following the first stage of execution will \emph{not}
automatically appear on the web interface.  To see the latest files,
we need to re-run stage one.  This cannot be done from within the \MOC
server page itself, but we have programmed the server to regenerate
pages whenever the `\cd{!}' character is appended to the URI.  The
directory browser pages feature a `Regenerate' button at the bottom
which will bring them up to date by running the publish stage code
again, and caching the result for subsequent requests.

@@

Currently, the \MOC implementation is not capable of reading
and writing \cd{code} values to a file.  This means that
generated server pages (after the publish stage) must already be
in the server memory when it starts.  This is an unfortunate
limitation, as it means we must restart the entire web server to
re-publish any particular page!

@@

\section{Performance}
\label{sec:perf}
One of the benefits of a \emph{compiled} implementation of staged
programming---even byte-code compiled, like \MOC---is that it permits
meaningful measurement.  Here, we describe the performance
characteristics of the prototype, focusing in particular on the
benefits of staging various web services.

The single most important metric for evaluating web server performance
is \textit{throughput:} the number of requests successfully answered
per unit of time.  To establish a baseline, we first tested the
throughput of the Caml prototype HTTP server delivering chunks of
static data of various sizes, up to 64k bytes.
%%
Measurements were taken on an otherwise idle 1.8GHz Intel
Xeon\texttrademark{} workstation\footnote{with 512kB cache, 768MB RAM,
  and Ultra160 SCSI} running Linux 2.4.  We used \texttt{ab},
the Apache HTTP server benchmarking tool, to issue requests
from 8 threads simultaneously for 30 seconds.\footnote{\texttt{ab}
  version 1.3d, available from \url{http://httpd.apache.org/}, 
  invoked like this:
  \texttt{ab -k -t 30 -c 8 }\textit{url}}

@@

The baseline results are shown in figure~\ref{fig:static}.  The series
marked “Caml file” was served by the \texttt{FileHandler} module.  For
the “Caml code” series, the same files were treated as \MOC Server
Pages, and thus turned into one big \texttt{puts} statement, to be
executed by the \texttt{CodeHandler}.  The only reason the code
wins in the beginning is that all code pages are already loaded into
the server’s memory on startup (to work around a limitation of the
prototype---see section~\ref{sec:impl}), but the \texttt{FileHandler}
must read files from disk each time.

Figure~\ref{fig:static} includes comparable results for PHP, the
popular server-side computation system.\footnote{libphp4.so
  (version 4.3.4) loaded into Apache 1.3} Here, we gave the same
static data files the extension \texttt{.php}, so that Apache
would treat them as PHP code, even though they have no
\cd{\Mquote{}code blocks\Munquote}.
%%
We omitted the results for Apache serving static files, because they
are way off scale: Apache handled an astounding 4,438 hits per second
for the 1k file, and 1,813 for the 64k file.  Apache is heavily
optimized for serving static files: apart from caching, it uses the
special \texttt{sendfile(2)} system call for zero-copy file transfer
from kernel space~\cite{tranter03sendfile}.  Although our prototype is
no match for Apache on static files, the overhead for interpreting
code seems no worse than that of PHP.
%%

@@

Finally, we look at the performance of staged and unstaged directory
browsing; see figure~\ref{fig:browse}.  For this task, we created
directories containing a fixed number of files with random data.  The
average file size was 32k.  The \emph{x}~axis shows the number of
files in the directory.  Besides the staging factor, two
implementation languages are compared: we coded the same functionality
in PHP.  The `staged PHP' version must be run first from the command
line; this outputs another PHP script which is then run by the server.

@@

Directory browsing was the most realistic of the examples, and the
benefit of staging is crystal clear.  In browsing a directory of 64
files, the unstaged programs barely answered 18 requests per second.
They would certainly succumb to the Slashdot effect!

This is exactly the kind of page where the real work needs to be done
in advance.  But that doesn't mean it needs to be a completely static
page, either.  By carefully staging the computation, we gathered the
file information in advance, yet still filtered and sorted the results
on demand.  The staged \MOC directory browser answered more than 550
requests (for the same 64-file directory) per second.

The degradation in performance of the staged PHP version is most
likely due to the fact that the stage two script must be \emph{parsed}
on each request; the size of the script is proportional to the number
of files in the directory.  With \MOC, the code size is also
proportional to the number of files, but the values are already in
memory.

\section{Related and future work}
\label{sec:related}

Most of the \emph{server page} systems embed programs within web pages
using similar techniques; exemplars include JSP~\cite{sun-jsp},
ASP~\cite{microsoft-asp}, PHP~\cite{bakken04php},
MSP~\cite{elsman03web} (based on SML), and XCaml~\cite{baretta04xcaml}
(based on OCaml).  Ours appears to be the first such language with
explicit support for \emph{staging} the computation.  We inherited the
staging annotations of \MOC~\cite{calcagno03meta} and designed several
new kinds of code blocks based on them.

Unlike more ad-hoc staging methods, \MOC server pages guarantee the
type safety for generated code up front.  We do not, however, make any
guarantees about the validity of the generated HTML.
\citet{elsman04typing}, \citet{wallace99haxml}, \citet{hosoya03xduce},
and \citet{ohl04xhtml} leverage ML-like type systems to validate
(X)HTML generators.  Integrating their ideas into \MOC server pages
may permit validation of the generated document as well.

Our server page language is defined by translation into \MOC.
Unfortunately, this means that error messages refer to the translated
code, not the original, embedded code.  Camlp4 is a flexible
pre-processor capable of modifying the concrete syntax of
OCaml~\cite{rauglaudre03camlp4} while maintaining usable error
messages.  Formulating the server page syntax with Camlp4 would likely
be an improvement over the current, ad-hoc translator.
